A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-07-19 rigid_bodies.tex
Finalization of Pressure Section and Changes in Results Section

Update of forces and torques is required only after the free translation and free rotation have been carried out. Each complete rotation is split as\cite{Miller2002}
\begin{equation}
\label{eq:splitting_rot}
e^{\Delta t i\!L_r} = \prod_{j=1}^n \left( e^{\frac{\delta t_j}{2} i\!L_3} e^{\frac{\delta t_j}{2} i\!L_2} e^{\delta t_j i\!L_1} e^{\frac{\delta t_j}{2} i\!L_2} e^{\frac{\delta t_j}{2} i\!L_3} \right),
\end{equation}
where $\sum_{j=1}^n {\delta t}_j = \Delta t$. This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step for accuracy improvement. Following Ref.~\onlinecite{Miller2002}, one could simply make $\delta t_j = \Delta t/n$ for all $j$. Alternatively, one could use a higher-order factorization scheme, such as the one due to Yoshida\cite{Yoshida1990} and Suzuki,\cite{Suzuki1991a, Suzuki1991b} in which $\delta t_j = \omega_j \Delta t$ and the weights $\omega_j$ are carefully chosen to determine the order of approximation. It is our intention in this paper to investigate how the performance of the integrator as a whole is affected by using such a higher-order scheme. For that, we will consider $n = 1$ with $\omega_1 = 1$ (zero-order) and $n = 7$ with $\omega_1 = \omega_7 = 0.784513610477560$, $\omega_2 = \omega_6 = 0.235573213359357$, $\omega_3 = \omega_5 = -1.17767998417887$, and $\omega_4 = 1 - \sum_{j \neq 4} \omega_j$ (sixth-order).

@@

Considering the system described above, with $\vt r^N$ and $\vt q^N$ representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
\begin{equation}
\label{eq:H_NVE}
\mathcal{H} = U(\vt r^N,\vt q^N) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right).
\end{equation}

@@

The usual way of computing pressure in molecular dynamics is based on.....

We now derive an expression for calculating the pressure of a rigid body system under periodic boundary conditions. For the sake of simplicity, we consider that the interaction potential is pairwise additive with respect to the atoms that form the rigid bodies. This enables us to follow the approach of Louwerse and Baerends.\cite{Louwerse2006} If multibody interactions are present, the approach of Thompson \textit{et al}.\cite{Thompson2009} is recommended. For a pairwise additive model, the potential energy of the $N$ rigid bodies whose centers of mass lie in a simulation box with periodic boundary conditions is
\begin{equation}
\label{eq:U_pairwise_pbc}
U({\vt r}, {\vt q}) = \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} u({\vt r}_{ij},{\vt q}_i,{\vt q}_j),
\end{equation}
where ${\vt r}_{ij} = {\vt r}_i - {\vt r}_j$ and
\begin{equation}
\label{eq:u_rigid_body_pair}
u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{{\vt R}_i}_k - {{\vt R}_j}_l\| \right).
\end{equation}

\begin{equation}
\label{eq:u_rigid_body_pair}
u({\vt r}_{ij},{\vt q}_i,{\vt q}_j) = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon \left( \|{\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l\| \right).
\end{equation}

In the equation above, ${{\vt R}_i}_k$ is the position of the $k$-th particle of body $i$ and $\upsilon(r)$ is the interaction potential between two atoms separated by a distance $r$. The dependence of this potential on the types of these two atoms is omitted for simplicity. From Sec.~\ref{sec:kinematics}, one can deduce that

The difference ${{\vt R}_i}_k - {{\vt R}_j}_l$ is given by
\[
{{\vt R}_i}_k - {{\vt R}_j}_l = {\vt r}_{ij} + {\mt A}({\vt q}_i){{\vt d}_i}_k - {\mt A}({\vt q}_j){{\vt d}_j}_l.
\]

The system pressure, considering a unit cell with constant volume $V$, is\cite{Louwerse2006,Tuckerman2010}
\[
P = \frac{(3N-3) k_B T}{3V} - \left\langle \diff{U}{V} \right\rangle.
\]

The first term in the expression above accounts for the ideal gas contribution. Applying the chain rule with Eq.~\ref{eq:U_pairwise_pbc}, the derivative of $U$ with respect to $V$, considering a cubic box with side length $L$, becomes
\[
%\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{u}{{\vt r}_{ij}} \cdot \diff{{\vt r}_{ij}}{L} + \right. \\ \left. \diff{u}{{\vt q}_i} \cdot \diff{{\vt q}_i}{L} + \diff{u}{{\vt q}_j} \cdot \diff{{\vt q}_j}{L}\right).
\diff{U}{V} = \diff{L}{V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \left( \diff{{\vt r}_{ij}}{L} \diff{u}{{\vt r}_{ij}} + \diff{{\vt q}_i}{L} \diff{u}{{\vt q}_i} + \diff{{\vt q}_j}{L} \diff{u}{{\vt q}_j} \right).
\]

Given that $L = V^\frac{1}{3}$ and that the bodies would neither deform nor rotate by a box volume change, we have that $\partial L / \partial V = 1/3L^2$, $\partial {\vt r}_{ij}/\partial L = \tr{\vt r}_{ij}/L$ (see Ref.~\onlinecite{Louwerse2006}) and $\partial {\vt q}_i / \partial L = \partial {\vt q}_j / \partial L = 0$. Therefore,
\[
\diff{U}{V} = \frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty} \tr{\vt r}_{ij} \diff{u}{{\vt r}_{ij}}.
\]

We can now use Eq.~\ref{eq:u_rigid_body_pair} to obtain the term inside the double summation above. By the chain rule, we have
\[
\diff{u}{{\vt r}_{ij}} = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \diff{\vt r}{\vt r_{ij}} \diff{\upsilon}{\vt r},
\]
where $\vt r = {{\vt R}_i}_k - {{\vt R}_j}_l$. Notice that, inside the double summation above, the first derivative is the identity matrix and the second one is $-{\vt f}_{ijkl}$, with this force being the action of the $l$-th particle of body $j$ over the $k$-th particle of body $i$. Therefore, the equation above results in $-\vt F_{ij}$, with this force being the resultant action of body $j$ over body $i$. Finally, we conclude that

\begin{equation}
\label{eq:pressure}
P = \frac{(N-1) k_B T}{V} + \frac{1}{3V} \left\langle \sum_{i=1}^{N-1} \sum_{j=i+1}^{\infty}{\vt F}_{ij} \cdot {\vt r}_{ij}  \right\rangle.
\end{equation}

This expression resembles the equation normally used to compute pressure in atomic systems, except that the internal virial is defined in terms of resultant interaction forces between rigid body pairs. Thus, an atom located outside the central box will contribute to this virial if it belongs to a body whose center lies in the box, and vice-versa.

\textbf{To do}: Write about minimum image convention considering rigid bodies instead of atoms and about how to compute the internal virial directly using the equation below:
\[
W = \sum_{i=1}^{N-1} \sum_{k=1}^{{n_p}_i} \sum_{j=i+1}^{\infty} \sum_{l=1}^{{n_p}_j} {\vt f}_{ijkl} \cdot {\vt r}_{ij}.
\]

@@

In this section, we analyze some performance aspects of the integration schemes when applied to a molecular system in the liquid state. Given its practical importance, we simulated 356 TIP3P\cite{Price2004} water molecules at a density of 999.56 $kg/m^3$. The intermolecular interactions were truncated at 9 {\AA}. The shifted-force (SF) method\cite{Allen1989,Toxvaerd_2011} was used to treat both the coulombic and the 12-6 Lennard-Jones (LJ) interactions.

@@

In order to evaluate the stability of the numerical solvers, we use a measure given by
\begin{equation}
\label{eq:performance}
D E =  \frac{1}{N_{steps}} \sum_{k=1}^{N_{steps}} \left| \frac{E_{t_k} - E_0}{E_0} \right|,
\end{equation}
from which it is possible to calculate the average deviation of the conserved quantity, i.e. total energy, from its initial value $E_0$ after a certain number of time steps $N_{steps}$. Both the instantaneous fluctuations of energy, given by the term inside the summation, and the long-time conservation of energy, $D E$, are monitored.

Recall that the method introduced here for the NVE dynamics is a mathematical reformulation of the NO$\rule{0.25cm}{0.15mm}$SQUISH algorithm introduced by Miller~\textit{et al.}\cite{Miller2002}, so the trajectories for both methods must coincide out of round-off errors. In Fig.~\ref{fig:miller1fs} we plot the instantaneous fluctuations of energy with $n_{ys} = 1 $ and a time step of 1 fs, where it can be seen that, in fact, the fluctuations for both methods are of similar magnitude. The simulation with the NO$\rule{0.25cm}{0.15mm}$SQUISH algorithm was conducted using the LAMMPS\cite{Plimpton1995} software package. For the purpose of comparison we applied the LJ model without shifting the forces, as it is not implemented in LAMMPS.

\begin{figure}[!h]
\centering
\includegraphics{millerourmd}
\captionof{figure}{Instantaneos fluctuations of energy for liquid water using Eq.~\ref{eq:trotter_splitting_NVE} (solid line) and Ref.\citenum{Miller2002} (dotted line) with $n = 1 $, $n_{ys} = 1$, $\Delta t = 1 fs$, $E_0 = -2799.87 \, kcal/mol $.}
\label{fig:miller1fs}
\end{figure}

In Table~\ref{table:nve} we compare the energy drift when using $n_{ys} = 1$ and $n_{ys} = 7$ in the rotational operator, Eq.~\ref{eq:ys_rot}. As expected, for a small time step of 1 fs, that further factorization is worthless. As the time step is increased, more accurate trajectories are achieved with $n_{ys} = 7 $. Nevertheless, the improvement is only marginal and all trajectories are stables exhibiting a small long-time energy drift. Thus, we proceed to analyze the NVT dynamics considering $n_{ys} = 1$.

\begin{table}[h]
\setlength{\tabcolsep}{7pt}
\caption{Average deviation of the conserved quantity ($DE$) from liquid water NVE simulations with $n_{ys}=1$ and $n_{ys}=7$  }
\centering % centering table
\begin{tabular}{| c  c  c |}  
\hline
&  \multicolumn{2}{c|}{ $D E$ }\\
\cline{2-3}
$\Delta t/fs$ &$n_{ys}=1$ &$n_{ys}=7$ \\
\hline % inserts single-line
 1 & 0.000055 & 0.000074\\
 2 & 0.00048 & 0.00031 \\
 3 & 0.0012 & 0.0011\\
 \hline
\end{tabular}
\label{table:nve}
\end{table}

It is not the purpose of this work to achieve an optimal factorization of the NVT schemes, but rather to show that a unique chain thermostat, while providing the ``proper'' sampling of the NVT ensemble, is capable of maintaining the average temperature at its specified value. In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\Delta t/4)$, we factorized those operators analogously to the scheme given by Eq.~\ref{eq:splitting_rot}. Recalling that in the NVT dynamics the conserved quantity is the extended energy $\mathcal{H}^\prime$, in Table~\ref{table:denvt} we show the results of Eq.~\ref{eq:performance} as well as the average temperatures obtained in each case. It can be seen that both integrators, represented by the Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:trotter_splitting_*}, systematically achieve the specified temperature, regardless of the time step size. In addition, the energy drift is, at its worst, around $1 \%$.

\begin{table}[h]
\setlength{\tabcolsep}{7pt}
\begin{threeparttable}
\caption{Average deviation of the conserved quantity ($DE$) and average temperatures from liquid water simulation with NVT integration \tnote{a}\tnote{b}} 
\label{table:denvt}

\centering % centering table
\begin{tabular}{|c c c c c |}  
\hline
& \multicolumn{2}{c}{Eq.~$\ref{eq:trotter_splitting_NHC}$} &\multicolumn{2}{c|}{Eq.~$\ref{eq:trotter_splitting_*}$} \\
\cline{2-5}
$\Delta t/fs$ &$D E$ &$T/K$ &$D E$ &$T/K$\\
\hline % inserts single-line
 1  & 0.000043 & 298.16  & 0.000052  & 298.15 \\

 2  & 0.00013 & 298.23  & 0.00017 & 298.06  \\

 3  & 0.00089 & 298.17  & 0.0013 & 298.12  \\

 \hline
\end{tabular}
\begin{tablenotes}
\item[a] The standard deviation of the mean temperature for all cases is $\sigma = 0.037$
\item[b] The smaller time step in the NHC and NHC* operators is 0.25 fs 
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

Checking the consistency between the equipartition of energy and the simulated energies serves as another approach for investigating the precision of a simulation. In fact, as stated by Eastwood~\textit{et al.}~\cite{Eastwood_2010}, it has been useful to improve methodologies related to the calculation of electrostatic\cite{Levitt_1988,Guenot_1992,Arnold_1994} and dispersive\cite{Sagui_1999} interactions, as well as to detect errors in thermostats\cite{Harvey_1998,Mor_2008} and barostats\cite{Feller_1995}. The truncation error associated to the numerical schemes is a potential source of error, and is the focus of the work of Eastwood~\textit{et al.}~\cite{Eastwood_2010}, which investigates a solvated protein and a temperature difference of approximately 5.5 $K$ is observed between the protein and the solvent using a time step of 2 fs.

In this paper, we analyze the partition of energy between the translational and rotational DOF, which are related to the temperature by the equipartition theorem as follows
\begin{align}
\label{eq:tras_equipartition}
2 \langle E_{kin}^{trans} \rangle = (3N - 3) k_b T
\end{align}

\begin{align}
\label{eq:rot_equipartition}
2 \langle E_{kin}^{rot} \rangle = 3N k_b T
\end{align} 
In these equations, $N$ is the number of rigid bodies. In NVE MD simulations, conservation of total linear momentum, $ \boldsymbol{P}= cte $, results in the loss of three translational DOF : $ 3N-3 $. For a cubic box with PBC there is no rotational symmetry ~\cite{Frenkel_2013,Kuzkin_2014}, with which the total angular momentum is not conserved, giving a number of rotational DOF of $3N$.  We calculate the average energy per DOF $ \langle E_{cin} \rangle_{DOF} = \langle E_{cin} \rangle /N_{DOF}$ and then we calculate both the translational and rotational energies from equipartition as follows

@@

\begin{table}[h]
\centering % centering table
\begin{threeparttable}
\caption{Partition of kinetic energy from NVE simulation of 903 water molecules with $r_c = 9 $ {\AA} and SF-LJ\tnote{a}} %($5415 grados de libertad$)}
%\begin{adjustbox}{width=1\textwidth}
\begin{tabular}{| l c  c  c  c  c  c |}  
\hline
 & & & \multicolumn{2}{c}{Equipartition} & \multicolumn{2}{c|}{Simulated}\\
  & & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c|}{partition}\\
\cline{4-7}
$\Delta t/fs$  &$T_{sim}/K$  &$E_{total}$ &$E_{trans}$ &$E_{rot}$ &$E_{trans}$ &$E_{rot}$ \\
\hline % inserts single-line
 0.5 & 305.20  & 1642.11  & 820.60  & 821.06  & 820.60 & 821.51  \\
  1  & 304.92  & 1640.57  & 819.83  & 820.74  & 820.63 & 819.94  \\
  2  & 304.20  & 1636.74  & 817.92  & 818.82  & 820.56 & 816.18 \\
  3  & 302.97  & 1630.07  & 814.59  & 815.48  & 820.47 & 809.60  \\
  \hline
 \end{tabular}
 \begin{tablenotes}
\item[a] Energies are in $kcal/mol$.
\end{tablenotes}
\label{table:water_partition_LJ-SF}
\end{threeparttable}
%\end{adjustbox}
\end{table}

>>>>

Update of forces and torques is required only after the free translation and free rotation have been carried out. Each complete rotation is split as\cite{Miller2002}
\begin{equation}
\label{eq:splitting_rot}
e^{\Delta t i\!L_r} = \prod_{j=1}^n \left( e^{\frac{\delta t_j}{2} i\!L_3} e^{\frac{\delta t_j}{2} i\!L_2} e^{\delta t_j i\!L_1} e^{\frac{\delta t_j}{2} i\!L_2} e^{\frac{\delta t_j}{2} i\!L_3} \right),
\end{equation}
where $\sum_{j=1}^n {\delta t}_j = \Delta t$. This means that free rotation, which is computationally inexpensive for not involving force calculations, can be carried out with a smaller time step for accuracy improvement. Following Miller \textit{et al}.,\cite{Miller2002} one could simply make $\delta t_j = \Delta t/n$ for all $j$. Alternatively, one could use a higher-order factorization scheme, such as the one due to Yoshida\cite{Yoshida1990} and Suzuki,\cite{Suzuki1991a, Suzuki1991b} in which $\delta t_j = \omega_j \Delta t$ and the weights $\omega_j$ are carefully chosen to determine the order of approximation. It is our intention in this paper to investigate how the performance of the integrator as a whole is affected by using such a higher-order scheme. For that, we will consider $n = 1$ with $\omega_1 = 1$ (zero-order) and $n = 7$ with $\omega_1 = \omega_7 = 0.784513610477560$, $\omega_2 = \omega_6 = 0.235573213359357$, $\omega_3 = \omega_5 = -1.17767998417887$, and $\omega_4 = 1 - \sum_{j \neq 4} \omega_j$ (sixth-order).

@@

Considering the system described above, with $\vt r$ and $\vt q$ representing the combined sets of all coordinates and quaternion components, we define the Hamiltonian
\begin{equation}
\label{eq:H_NVE}
\mathcal{H} = U(\vt r,\vt q) + \sum_{i=1}^N \left(\frac{\tr{\vt p}_i{\vt p}_i}{2m_i} + \frac{\tr{\vt \pi}_i {\mt B}_i {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i}{8}\right).
\end{equation}

@@

The pressure of a system with constant NVT, for which $\sum_{i=1}^N{\vt F}_i = 0$, is calculated by\cite{Louwerse2006, Tuckerman2001}
\[
P = \frac{(N-1) k_B T + \left\langle W \right\rangle}{V},
\]
where $W$ is the internal virial of the system, defined as $W = - V \frac{dU}{dV}$. For simplicity, we consider a system of rigid bodies under periodic boundary conditions whose potential energy function is pairwise additive with respect to individual atoms, so that the approach of Louwerse and Baerends\cite{Louwerse2006} can be applied (otherwise, one can employ the approach of Thompson \textit{et al.}\cite{Thompson2009}). Due to the rigidity of the bodies, interactions amongst companion atoms (i.e., atoms belonging to the same body) are irrelevant. Assuming that the minimum image convention\cite{Allen1989} applies for entire bodies, meaning that the simulation box is large enough to preclude any atom from having their own images or images of its companion atoms within its cutoff radius, the potential energy per unit box is given by
\begin{equation}
\label{eq:U_pairwise_pbc}
U({\vt r}, {\vt q}) = \sum_{i=1}^{N-1} \sum_{j=i+1}^N \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \upsilon(r_{ijkl}),
\end{equation}
where $\upsilon(r)$ is a pair interaction potential and $r_{ijkl}$ is the distance between the $k$-th atom of body $i$ to its nearest image of the $l$-th atom of body $j$. Such distance is the norm of a vector ${\vt r}_{ijkl} = \hat{\vt r}_{ij} + \tr{[{\mt A}({\vt q}_i)]}{{\vt d}_k}_i - \tr{[{\mt A}({\vt q}_j)]}{{\vt d}_l}_k$, where $\hat{\vt r}_{ij}$ is the minimum-image vector pointing from the center of mass of body $j$ to that of body $i$. We note that $U$ does not depend explicitly on $V$, which makes ${\partial U}/{\partial V} = 0$. If the atoms were free to move individually, an isotropic change in volume would displace all atoms and we would have ${\partial {\vt r}_{ijkl}}/{\partial V} = (3V)^{-1} \tr{\vt r}_{ijkl}$.\cite{Louwerse2006} In this event, the atomic internal virial would be computed by
\begin{equation}
\label{eq:virial_atoms}
W_\text{a} = \frac{1}{3} \sum_{i=1}^{N-1} \sum_{j=i+1}^N \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \tr{\vt r}_{ijkl} {\vt f}_{ijkl},
\end{equation}
where $\vt f_{ijkl} = -{\partial \upsilon}/{\partial \vt r_{ijkl}}$ is the minimum-image force exerted by the $l$-th atom of body $j$ on the $k$-th atom of body $i$. In the case of rigid bodies, considering that an isotropic change in volume would neither deform nor rotate them, but only promote center-of-mass displacements, we have that ${\partial \hat{\vt r}_{ij}}/{\partial V} = (3V)^{-1} \tr{\hat{\vt r}}_{ij}$ and ${\partial {\vt q}_i}/{\partial V} = {\partial {\vt q}_j}/{\partial V} = \tr{\vt 0}$. Therefore, by applying the chain rule to Eq.~\ref{eq:U_pairwise_pbc} and realizing that $\partial \vt r_{ijkl}/\partial \hat{\vt r}_{ij} = \vt 1_3$, we obtain
\[
\label{eq:dUdV_rigid_bodies}
\frac{dU}{dV} = -\frac{1}{3V} \sum_{i=1}^{N-1} \sum_{j=i+1}^N \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \tr{\hat{\vt r}}_{ij} \vt f_{ijkl}.
\]

There are two convenient ways of employing this result to compute the internal virial of a system of rigid bodies. On one hand, we can explore the fact that ${\hat{\vt r}}_{ij}$ does not depend on the indexes $k$ and $l$. On the other hand, we can replace $\hat{\vt r}_{ij} = {\vt r}_{ijkl} - {{\vt \delta}_k}_i + {{\vt \delta}_l}_k$ and make use of the fact that $\vt f_{ijkl} = -\vt f_{jilk}$. Finally, we have
\begin{subequations}
	\label{eq:virial_rigid_bodies}
	\begin{align}
	W &= \frac{1}{3} \sum_{i=1}^{N-1} \sum_{j=i+1}^N \tr{\hat{\vt r}}_{ij} \vt F_{ij} \\
	  &= W_\text{a} - \frac{1}{3} \sum_{i=1}^N \sum_{k=1}^{{n_p}_i} {\tr{\vt \delta}_k}_i {{\vt f}_k}_i,
	\end{align}
\end{subequations}
where $W_\text{a}$ is the atomic virial given by Eq.~\ref{eq:virial_atoms}, while $\vt F_{ij}$ is the total action of body $j$ (and its images) over body $i$ and $\vt f_{ik}$ is the resultant force exerted on the $k$-th atom of body $i$, respectively given by
\
\[
\vt F_{ij} = \sum_{k=1}^{{n_p}_i} \sum_{l=1}^{{n_p}_j} \vt f_{ijkl} \quad \text{and} \quad \vt f_{ik} = \sum_{\substack{j=1\\j \neq i}}^N \sum_{l=1}^{{n_p}_j} \vt f_{ijkl}.
\]

@@

In this section, we analyze some performance aspects of the integration schemes when applied to a molecular system in the liquid state. For this, we simulated a system containing 356 TIP3P water molecules\cite{Price2004} at a density of 999.56 $kg/m^3$. The intermolecular interactions were truncated at 9 \AA. The shifted-force (SF) method\cite{Allen1989, Fennell2006, Toxvaerd_2011} was used to treat both the Coulombic and the 12-6 Lennard-Jones (LJ) interactions.

@@

In order to evaluate the stability of the numerical integrators, we employ a measure given by\cite{Tuckerman2010}
\begin{equation}
\label{eq:performance}
D\!E =  \frac{1}{N_\text{steps}} \sum_{k=1}^{N_\text{steps}} \left| \frac{E_{t_k} - E_0}{E_0} \right|,
\end{equation}
from which it is possible to quantify the average deviation of the supposedly conserved quantity $E$ (either $\mathcal{H}$ of Eq.~\ref{eq:H_NVE} or $H$ of Eq.~\ref{eq:H_nvt}) from its initial value $E_0$ after a certain number of time steps $N_\text{steps}$. Both the instantaneous fluctuations, given by the term inside the summation, and the long-time conservation of energy, $D\!E$, are monitored.

Recall that, when the splitting scheme of Eq.~\ref{eq:splitting_rot} is executed with $\delta t_j = \Delta t/n$ for all $j$, the method introduced here for the NVE dynamics is just a mathematical reformulation of the algorithm of Miller \textit{et al}.\cite{Miller2002} Therefore, short trajectories obtained with both methods should coincide up to round-off errors. In Fig.~\ref{fig:miller1fs}, we plot the instantaneous fluctuations of total energy obtained with $n = 1$ and a time step of 1~fs using both methods. The simulation with the algorithm of Ref.~\onlinecite{Miller2002} was performed using the LAMMPS software package.\cite{Plimpton1995} For the purpose of comparison, we did not carry out force shifting for the Lennard-Jones model, as this is not implemented in LAMMPS. It can be seen in Fig.~\ref{fig:miller1fs} that, in fact, the fluctuations of both methods are of similar magnitudes.

\begin{figure}
	\includegraphics{millerourmd}
	\caption{Instantaneos fluctuations of energy for liquid water using Eq.~\ref{eq:trotter_splitting_NVE} (solid line) and the LAMMPS implementation of the algorithm of Ref.~\onlinecite{Miller2002} (dotted line) with $n = 1$, $\Delta t = 1$ fs, and $E_0 = -2799.87$ kcal/mol.}
	\label{fig:miller1fs}
\end{figure}

In Table~\ref{table:nve}, we compare the energy drifts observed when the rotational operator in Eq.~\ref{eq:splitting_rot} is split using either $n = 1$ or the sixth-order Yoshida-Suzuki factorization scheme ($n = 7$). As observed, for a small time steps of 1~fs, increasing the order of factorization does not provide any improvement. As the time step size increases, trajectories with higher accuracy are achieved with $n = 7$. Nevertheless, the improvement is only marginal and all trajectories are stable, exhibiting a small long-time energy drift. Thus, we proceed by considering $n = 1$ in the analysis of the NVT dynamics.

\begin{table}
	\caption{Average total-energy deviations for liquid water in NVE simulations for Eq.~\ref{eq:splitting_rot} with either $n = 1$ or $n = 7$ with Yoshida-Suzuki factorization.}
	\label{table:nve}
	\begin{ruledtabular}
		\begin{tabular}{ccc}
			 & \multicolumn{2}{c}{ $D\!E$ } \\
			 \cline{2-3}
			 $\Delta t$/fs & $n = 1$ & $n = 7$ \\
			 \hline
			 1.0 & 0.000055 & 0.000074 \\
			 2.0 & 0.00048  & 0.00031  \\
			 3.0 & 0.0012   & 0.0011   \\
		\end{tabular}
	\end{ruledtabular}
\end{table}

It is not the purpose of this work to propose an optimal factorization scheme for the NVT dynamics, but rather to show that a unique thermostat chain is capable of maintaining the average temperature at its specified value while providing the proper sampling of the canonical ensemble. \textbf{In order to use a constant value for the smaller time step in the NHC and NHC* operators $(\Delta t/4)$, we factorized those operators analogously to the scheme given by Eq.~\ref{eq:splitting_rot}}. Recalling that in the NVT dynamics the conserved quantity is the extended energy $H$, in Table~\ref{table:denvt} we show the results of Eq.~\ref{eq:performance} as well as the average temperatures obtained in each case. It can be seen that both integrators, represented by the Eq.~\ref{eq:trotter_splitting_NHC} and Eq.~\ref{eq:modified_splitting}, systematically achieve the specified temperature, regardless of the time step size. In addition, the energy drift is, at its worst, around $1 \%$.

\begin{table}
	\begin{threeparttable}
		\caption{Average deviation of the conserved quantity ($DE$) and average temperatures from liquid water simulation with NVT integration \tnote{a}\tnote{b}}
		\label{table:denvt}
		\begin{ruledtabular}
			\begin{tabular}{ccccc}
				& \multicolumn{2}{c}{Eq.~\ref{eq:trotter_splitting_NHC}} & \multicolumn{2}{c}{Eq.~\ref{eq:modified_splitting}} \\
				\cline{2-5}
				$\Delta t$/fs & $D\!E$ & $T$/K & $D\!E$ & $T$/K \\
				\hline
				1.0 & 0.000043 & 298.16  & 0.000052  & 298.15 \\
				2.0 & 0.00013  & 298.23  & 0.00017   & 298.06 \\
				3.0 & 0.00089  & 298.17  & 0.0013    & 298.12
			\end{tabular}
		\end{ruledtabular}
		\begin{tablenotes}
			\item[a] The standard deviation of the mean temperature for all cases is $\sigma = 0.037$.
			\item[b] The smaller time step in the NHC and NHC* operators is 0.25 fs.
		\end{tablenotes}
	\end{threeparttable}
\end{table}

@@

Checking the consistency between the equipartition of energy and the simulated energies serves as another approach for investigating the precision of a simulation. In fact, as stated by Eastwood \textit{et al}.,\cite{Eastwood_2010} this kind of checking has been useful to improve methodologies related to the calculation of  electrostatic\cite{Levitt_1988, Guenot_1992, Arnold_1994} and dispersive\cite{Sagui_1999} interactions, as well as to detect errors in thermostats\cite{Harvey_1998, Mor_2008} and barostats.\cite{Feller_1995} The truncation error associated to the numerical schemes is a potential source of error, and is the focus of the work of Eastwood \textit{et al}.,\cite{Eastwood_2010} which investigates a solvated protein and a temperature difference of approximately 5.5~K is observed between the protein and the solvent using a time step of 2~fs.

In this paper, we analyze the partition of energy between the translational and rotational degrees of freedom, which are related to the temperature by the equipartition theorem as follows
\begin{subequations}
\begin{align}
\label{eq:tras_equipartition}
2 \langle E_\text{trans} \rangle = (3N - 3) k_b T
\end{align}
\begin{align}
\label{eq:rot_equipartition}
2 \langle E_\text{rot} \rangle = 3N k_b T
\end{align} 
\end{subequations}

In these equations, $N$ is the number of rigid bodies. In microcanonical MD simulations, conservation of total linear momentum $\boldsymbol{P}$ results in the loss of three translational degrees of freedom, which is the reason why the factor $3N-3$ appears in Eq.~\ref{eq:tras_equipartition}. For a cubic box with periodic boundary conditions, there is no rotational symmetry,\cite{Frenkel_2013, Kuzkin_2014} with which the total angular momentum is not conserved, giving a number of $3N$ rotational degrees of freedom. We calculate the average energy per DOF $\langle E_{cin} \rangle_{DOF} = \langle E_{cin} \rangle /N_{DOF}$ and then we calculate both the translational and rotational energies from equipartition as follows

@@

\begin{table}
	\begin{threeparttable}
		\caption{Kinetic energy partition in NVE simulations of 903 water molecules with $r_c = 9$ {\AA} and a shifted-force Lennard-Jones potential.\tnote{a}} %($5415 grados de libertad$)}
		\label{table:water_partition_LJ-SF}
		\begin{ruledtabular}
			\begin{tabular}{ccccccc}
				& & & \multicolumn{2}{c}{Equipartitioned} & \multicolumn{2}{c}{Simulated}\\
%				& & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c}{partition}\\
				\cline{4-7}
				$\Delta t$/fs & $T_\text{sim}$/K & $\langle E_\text{kin}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ \\
				\hline
				0.5 & 305.20  & 1642.11  & 820.60  & 821.06  & 820.60 & 821.51  \\
				1.0 & 304.92  & 1640.57  & 819.83  & 820.74  & 820.63 & 819.94  \\
				2.0 & 304.20  & 1636.74  & 817.92  & 818.82  & 820.56 & 816.18  \\
				3.0 & 302.97  & 1630.07  & 814.59  & 815.48  & 820.47 & 809.60
			\end{tabular}
		\end{ruledtabular}
		\begin{tablenotes}
			\item[a] Energies are in kcal/mol.
		\end{tablenotes}
	\end{threeparttable}
\end{table}

\begin{table}
	\begin{threeparttable}
		\caption{Kinetic energy partition in NVE simulations of 903 water molecules with $r_c = 13$ {\AA} and a shifted-force Lennard-Jones potential.\tnote{a}} %($5415 grados de libertad$)}
		\label{table:water_partition_13A_LJ-SF}
		\begin{ruledtabular}
			\begin{tabular}{ccccccc}
				& & & \multicolumn{2}{c}{Equipartitioned} & \multicolumn{2}{c}{Simulated}\\
%				& & & \multicolumn{2}{c}{of energy} & \multicolumn{2}{c}{partition}\\
				\cline{4-7}
				$\Delta t$/fs & $T_\text{sim}$/K & $\langle E_\text{kin}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ & $\langle E_\text{trans}\rangle$ & $\langle E_\text{rot}\rangle$ \\
				\hline
				0.5 & 307.17 & 1652.71  & 825.90  & 826.81  & 825.96 & 826.75  \\
				1.0 & 306.98 & 1651.68  & 825.38  & 826.30  & 826.09 & 825.59  \\
				2.0 & 306.04 & 1646.60  & 822.85  & 823.75  & 825.57 & 821.03  \\
				3.0 & 304.79 & 1639.89  & 819.49  & 820.40  & 825.63 & 814.26
			\end{tabular}
		\end{ruledtabular}
		\begin{tablenotes}
			\item[a] Energies are in kcal/mol.
		\end{tablenotes}
	\end{threeparttable}
\end{table}
